---
title: "Step1 update and download met files"
author: "Jackie"
date: "2025-04-01"
output: html_document
---

```{r setup_install and load packages, include=FALSE}
#install.packages("SWTools")
library(SWTools)
library(readr)
library(DT)

```

# Workflow for analysing climate data creating decile tables and forecasting yields.
* Series of R markdown. Best to run these in order, but once you have historical yields step 2 and 3 they don’t need to be repeated to get a yield forecast.  
**Step 1** 
 *	Download patch point met file, tidy up and saved to directory. User to specify the site number.  
 *  This is super glitchy. If you run it as a rmarkdown the new data won' be updated so run this code chunck by code chunck.
 
**Step 2** 
*	User defined site, and ‘season’.
*	Create plot sum of rain, per year, per season for site selected
*	Create historical files 1. Decile table for GS. 2. Decile table summer season.
*	File for each year with sum of rain per year and season type.  
**Step 3**
*	User defined site creates yield potential for decile years, using baseline and frontier approach.
*	Creates plot  
**Step 4**
*	User defined site creates yield forecast for decile years, using baseline and frontier approach for defined dates
*	Creates plot  


## Choose what site you want to download met file.


```{r list of sites, echo=FALSE}
list_of_sites <- data.frame(
    Name = c("Karoonda", "Wharminda", "Copeville", "Walpeup" , "Bute", "Balaklava", "Annuello", "Crystal Brook"), ### add more here when you have the info
    Number = c("25006", "18113", "25003", "76065", "21012", "21002", "76000", "21016"),
    Latitude = c("-35.09", "-33.97", "-34.80", "-35.14",  "-33.86", "-34.14", "-34.85",   "-33.35"),
    Longitude = c("139.90", "136.25", "139.85", "142.02", "138.01", "138.42", "142.78", "138.21"),
    State = c("SA", "SA", "SA", "VIC","SA", "SA", "VIC", "SA")
    )

DT::datatable(list_of_sites)


 
```


## Select the site number
* create a new variable in the below code chuck that will be used for the analysis
* Shiny application would be nice but then I have problem saving the output.

```{r sites_selction, echo=FALSE}
 site_selected <- "76065"
print(site_selected)
```



## Download or update SILO patchpoint using SWTool package
https://www.rdocumentation.org/packages/SWTools/versions/1.1.0/topics/SILODownload


* This package download the file/s and saves it as text file with the name of the station
* Here I have the files saved to  "H:/Output-2/Analysis/Scripts/Jackie/Downloaded_files"
* Last day of data, in the format "YYYYMMDD". Will default to yesterday if not specified


* If you need to change where this saved then change in the below code chuck.

```{r Download_parameters, eval=FALSE, include=FALSE}

Site_number <- site_selected
email_jackie <-"jackie.ouzman@csiro.au"
START <- "19000101"
#FINISH <- "20050203"
Files_saved <- "H:/Output-1/0.Site-info/Climate/Downloaded_files"


## this will download patchpoint data and save it to file directory that is specified
SILODownload(
  SiteList= Site_number,
  #SiteList = c("25006" ,"18113", "25003" ,"76065" ,"21012" ,"21002", "76000"),
  username = email_jackie,
  password = "gui",
  path = Files_saved,
  startdate = START,
  enddate = ,
  ssl = FALSE
)

```

# Get the downloaded climate file 
* This is patch point data that matches the site number
* Ensure the directories are correct

```{r clean_files, echo=FALSE, message=FALSE, warning=FALSE}
Files_saved <- "H:/Output-1/0.Site-info/Climate/Downloaded_files"

df <- read_table(
  paste0(Files_saved,"/", site_selected,".txt"),
  col_types =   cols(`Date2` = col_date(format = "%d-%m-%Y"),
  `Day` =   col_number()  ,
  `Radn` = col_number(),
  `T.Max` = col_number(),
  `T.Min` = col_number(),
  `Rain` = col_number(),
  `Evap` = col_number(),
  `VP` = col_number()),
        skip = 51
      )
#head(df)

#remove first row
df <- df %>% dplyr::filter( Date != "(yyyymmdd)")


 # select clms

df <- df %>%  dplyr::select(Date  ,
                                     Day ,
                                     Date2 ,
                                     T.Max ,
                                     T.Min ,
                                     Rain ,
                                     Evap   ,
                                     Radn )
 
df <- df %>%
  dplyr::rename(day_of_year = Day,
         Date_string = Date,
         Date = Date2
         ) %>%
  dplyr::mutate(year = lubridate::year(Date),
         month =lubridate::month(Date),
         day_of_month = lubridate::day(Date),
         month_name = lubridate::month(Date, label = TRUE)
  )


df_date_download <- read_csv(
  paste0(Files_saved, "/",site_selected,".txt"),
  col_names = FALSE, skip = 15) %>% dplyr::slice_head()


df_date_download$X1 <- gsub('[^[:alnum:] ]', '', df_date_download$X1)
df_date_download$X1 <- gsub(" Extracted from SILO BoM Only dataset on " , '', df_date_download$X1)







df <- df %>% dplyr::mutate(Number = site_selected)
df<- dplyr::left_join(df,list_of_sites)

write_csv(df,
          file = paste0(Files_saved,"/Neater_file_",site_selected,".csv"))

DT::datatable(head(df))



```

## List of files and date last saved in directory
```{r files in directory, echo=FALSE, message=FALSE, warning=FALSE}
Files_saved <- "H:/Output-1/0.Site-info/Climate/Downloaded_files"
files <- list.files(path = Files_saved,full.names = TRUE, pattern = ".csv")

file_info <- file.info(files)
file_info$file_name <- rownames(file_info)
file_info <- file_info %>%
  dplyr::select(atime)
#names(file_info)
DT::datatable(file_info)
```

# Just some checks is the script working?
1. What have you just downloaded? 
2. Does it match what was written to the download folder

## what is the last date in the data just downloaded?
```{r last date in the df, echo=FALSE, message=FALSE, warning=FALSE}
max(df$Date)

```
## what is the last date in the file just downloaded?

```{r site in the dowload folder, echo=FALSE, message=FALSE, warning=FALSE}

file_path_input_data<-file.path("H:","Output-1", "0.Site-info", "Climate", "Downloaded_files")


df_in_download_folder <- read.csv(paste0(file_path_input_data, "/Neater_file_", site_selected,  ".csv")) 


print( max(df_in_download_folder$Date))

```
## what is the site in the file just downloaded?

```{r last date in the dowload folder, echo=FALSE, message=FALSE, warning=FALSE}

print( unique(df_in_download_folder$Name))
       
```