---
title: "Step 2 creating decile tables basic"
author: "Jackie"
date: "2025-04-02"
output: html_document
---

```{r load_libary, include=FALSE}
library(ggplot2)
library(readxl)
library(tidyverse)
library(lubridate)
library(data.table)
library(stringr)
```

# Workflow for analysing climate data creating decile tables and forecasting yields.
* Series of R markdown. Best to run these in order, but once you have historical yields step 2 and 3 they don’t need to be repeated to get a yield forecast.  
**Step 1** 
 *	Download patch point met file, tidy up and saved to directory. User to specify the site number.  
**Step 2** 
*	User defined site, and ‘season’.
*	Create plot sum of rain, per year, per season for site selected
*	Create historical files 1. Decile table for GS. 2. Decile table summer season.
*	File for each year with sum of rain per year and season type.  
**Step 3**
*	User defined site creates yield potential for decile years, using baseline and frontier approach.
*	Creates plot  
**Step 4**
*	User defined site creates yield forecast for decile years, using baseline and frontier approach for defined dates
*	Creates plot  

## How to defined the start and end of growing season
* The default is 1/4 (1st April) to 15/10 (15th Oct)
* If you want to change this modified the below code chunk.

```{r define start and end of GS, echo=FALSE}


Day_start_GS_rainfall <- 1
Month_start_GS_rainfall <- 4
    

Day_end_GS_rainfall <- 15
Month_end_GS_rainfall <- 10
      

```


## How to defined the start and end of summer months
* The default is 1/11 (1st Nov) to 31/03 (31th March)
* If you want to change this modified the below code chunk.

```{r define start and end of summer months, echo=FALSE}



### summer dates after GS
Day_start_Summer_rainfall_after_GS <- 1
Month_start_summer_rainfall_after_GS <- 11

Day_end_summer_rainfall_after_GS <- 31 
Month_end_summer_rainfall_after_GS <- 12

### summer dates before GS
Day_start_Summer_rainfall_before_GS <- 1
Month_start_summer_rainfall_before_GS <- 1

Day_end_summer_rainfall_before_GS <- 1 # if you want this to be the 31 /3 you need to add a day ie 1/4
Month_end_summer_rainfall_before_GS <- 4
      

```

## Choose what site you want to create decile table for.


```{r list of sites, echo=FALSE}
list_of_sites <- data.frame(
    Name = c("Karoonda", "Wharminda", "Copeville", "Walpeup" , "Bute", "Balaklava", "Annuello", "Crystal Brook"), ### add more here when you have the info
    Number = c("25006", "18113", "25003", "76065", "21012", "21002", "76000", "21016"),
    Latitude = c("-35.09", "-33.97", "-34.80", "-35.14",  "-33.86", "-34.14", "-34.85",   "-33.35"),
    Longitude = c("139.90", "136.25", "139.85", "142.02", "138.01", "138.42", "142.78", "138.21"),
    State = c("SA", "SA", "SA", "VIC","SA", "SA", "VIC", "SA")
    )

DT::datatable(list_of_sites)
```


## List of files in step 1
* Looking in this folder.
* If it is in another location change the bellow code chuck.(file_path_input_data)
* "//fs1-cbr.nexus.csiro.au/{af-sandysoils-ii}/work/Output-2/Analysis/Scripts/Jackie/Downloaded_files"

```{r list of sites with decile tables, echo=FALSE}

 file_path_input_data<-file.path("H:","Output-1", "0.Site-info", "Climate", "Downloaded_files")

#"H:/Output-1/0.Site-info/Climate/Downloaded_files"

list_of_file <- list.files(path = file_path_input_data, pattern = ".csv")
list_of_file <- as.data.frame(list_of_file)
DT::datatable(list_of_file)
 
```

## Select the site number
* create a new variable in the below code chuck that will be used for the analysis
* Shiny application would be nice but then I have problem saving the output.

```{r sites_selction, echo=FALSE}
 site_selected <- "21016"
print(site_selected)
```
## Download the **Neater met file**  

```{r sites_selction_download, echo=FALSE}

path <-  file.path("H:","Output-1", "0.Site-info", "Climate", "Downloaded_files")
          
df <- read.csv(paste0(path, "/Neater_file_", site_selected,  ".csv")) 


## add a clm with two dates marked as start and end of GS
df <- df %>% mutate(
  start_end_GS_date = case_when(
    month == Month_start_GS_rainfall & day_of_month == Day_start_GS_rainfall ~ "start_gs",
    month == Month_end_GS_rainfall & day_of_month == Day_end_GS_rainfall ~ "end_gs",
    TRUE ~ NA))

### fill the blanks

df <- df %>% fill(start_end_GS_date) %>%
  mutate(
    season_GS = case_when(
      start_end_GS_date == "start_gs" ~ "gs",
      TRUE ~ "other"))

df <- df %>% select(-start_end_GS_date)


## add a clm with two dates marked as start and end of Summer

df <- df %>% mutate(
  start_end_summer_date = case_when(
    #before GS
    month == Month_start_summer_rainfall_before_GS & 
      day_of_month == Day_start_Summer_rainfall_before_GS ~ "start_summer", 
    month == Month_end_summer_rainfall_before_GS & 
      day_of_month == Day_end_summer_rainfall_before_GS ~ "end_summer",
    
    #after GS
    month == Month_start_summer_rainfall_after_GS & 
      day_of_month == Day_start_Summer_rainfall_after_GS ~ "start_summer", 
    month == Month_end_summer_rainfall_after_GS & 
      day_of_month == Day_end_summer_rainfall_after_GS ~ "end_summer", 
    
        TRUE ~ NA))

### fill the blanks

df <- df %>% fill(start_end_summer_date) %>%
  mutate(
    season_summer = case_when(
      start_end_summer_date == "start_summer" ~ "summer",
      month == 12 & day_of_month == 31 ~ "summer",
      TRUE ~ "other"))

df <- df %>% select(-start_end_summer_date)


df <- df %>% mutate(
  season = (case_when(
    season_summer == "summer" ~ "summer",
    season_GS == "gs" ~ "gs",
    TRUE ~ "other"
  )))

      
``` 

#### Date range of file you have selected.
```{r sites_selction_date range, echo=FALSE}

#File start date and end date
paste("Start date in file is: ",
      min(df$Date),
      ". End date in file is: ",
      max(df$Date))
      
 
paste("Site Name (BOM): ", unique(df$Name),
      ". Site Number (BOM): ", unique(df$Number)
      )
           

GS_defined_as <- paste0(
  Day_start_GS_rainfall, "/", Month_start_GS_rainfall,
  " to ", Day_end_GS_rainfall, "/", Month_end_GS_rainfall)

summer_defined_as <- paste0(
  Day_start_Summer_rainfall_after_GS, "/", Month_start_summer_rainfall_after_GS,
  " to ", Day_end_summer_rainfall_before_GS, "/", Month_end_summer_rainfall_before_GS)

paste("summer month: ", summer_defined_as)
paste("GS season: ", GS_defined_as)

                
      
```  

### Do you want to use this full data set?
* The default is 1/1/1900 to 31/12/2025
* If you want to change it modify the below code chunk.

```{r subset date range, echo=FALSE}
df_all_date <- df #keep this

start <- "1900-01-01"
end <- "2025-12-31"

### Filter dataset based on years 
df <- df %>% filter(between(Date, as.Date(start), as.Date(end)))

```

## Plot rain for each season type

```{r plot rain by season type, echo=FALSE, message=FALSE, warning=FALSE}


plot1 <- df %>%
   filter(season!= "other") %>% 
    group_by(year, season) %>%
    summarise(sum_rain_season = sum(Rain, na.rm = TRUE)) %>%
    
    
    ggplot(aes(x = year, y = sum_rain_season, fill = season)) +
    geom_bar(stat = "identity") +
    labs(
      title = "Sum of rain per year",
      subtitle = paste0(
        "Station name: ",
        distinct(df, Name)
      ),
      
       caption = paste0("Years included : ",
                        year(min(df$Date)) ,
                        " to ",
                        year(max(df$Date)),
                       ". GS defined as: ", 
                         GS_defined_as),
      
      y = "sum of rain")
plot1

```

## GS decile table 

```{r decile table GS, echo=FALSE, message=FALSE, warning=FALSE}



  
  gs_rain <- df %>%
    filter(season!= "other") %>% 
    group_by(year, season) %>%
    summarise(sum_rain_season = sum(Rain, na.rm = TRUE)) %>%
    filter(season == "gs")
  
  ## 5a. GS ----------------------------------------------------------------------
#the percentiles for each year.

GS_decile_max_rain <- quantile(gs_rain$sum_rain_season, c(.1, .2, .3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0))
GS_deciles_names <- c("decile_1", "decile_2", "decile_3", "decile_4", "decile_5", "decile_6", "decile_7", "decile_8", "decile_9", "decile_10")

GS_decile_table <- data.frame(GS_decile_max_rain, GS_deciles_names, row.names = NULL)
GS_decile_table <- mutate(GS_decile_table, GS_decile_min_rain = lag(GS_decile_max_rain+0.1))


#add in the min value to table
GS_decile_table[1,3] <-  min(gs_rain$sum_rain_season)
GS_decile_table <- GS_decile_table %>%  select (GS_deciles_names, GS_decile_min_rain, GS_decile_max_rain)


 GS_decile_table <- GS_decile_table %>% 
   mutate(Station_name = unique(df$Name),
          Station_number = unique(df$Number),
           Years_included = paste0(
                                   year(min(df$Date)) ,  " to ", 
                                   year(max(df$Date))),
          GS_period =   GS_defined_as,
          summer_period =   summer_defined_as)
          
                         

GS_decile_table_df <- as.data.frame(GS_decile_table)
GS_decile_table_df <- GS_decile_table_df %>% mutate(across(where(is.numeric), round, 2))
#won't save because its reactive - I cant work it out

DT::datatable(GS_decile_table_df)

```

## Summer decile table 

```{r decile table Summer, echo=FALSE, message=FALSE, warning=FALSE}


  
  summer_rain <- df %>%
    filter(season!= "other") %>% 
    group_by(year, season) %>%
    summarise(sum_rain_season = sum(Rain, na.rm = TRUE)) %>%
    filter(season == "summer")
  
  ## 5b. Summer ----------------------------------------------------------------------
#the percentiles for each year.

summer_decile_max_rain <- quantile(summer_rain$sum_rain_season, c(.1, .2, .3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0))
summer_deciles_names <- c("decile_1", "decile_2", "decile_3", "decile_4", "decile_5", "decile_6", "decile_7", "decile_8", "decile_9", "decile_10")

summer_decile_table <- data.frame(summer_decile_max_rain, summer_deciles_names, row.names = NULL)
summer_decile_table <- mutate(summer_decile_table, summer_decile_min_rain = lag(summer_decile_max_rain+0.1))


#add in the min value to table
summer_decile_table[1,3] <-  min(summer_rain$sum_rain_season)
summer_decile_table <- summer_decile_table %>%  select (summer_deciles_names, summer_decile_min_rain, summer_decile_max_rain)


 summer_decile_table <- summer_decile_table %>% 
   mutate(Station_name = unique(df$Name),
          Station_number = unique(df$Number),
           Years_included = paste0(
                                   year(min(df$Date)) ,  " to ", 
                                   year(max(df$Date))),
          GS_period =   GS_defined_as,
          summer_period =   summer_defined_as
          )
          
                         

summer_decile_table_df <- as.data.frame(summer_decile_table)
summer_decile_table_df <- summer_decile_table_df %>% mutate(across(where(is.numeric), round, 2))
#won't save because its reactive - I cant work it out


  
DT::datatable(summer_decile_table_df)
```

### Assign decile to each year
* Create a dataframe with year, summer rain, GS rain, summer decile, winter decile.

```{r years rainfall sum decile, echo=FALSE, message=FALSE, warning=FALSE}

gs_rain <- gs_rain %>%
  mutate(
    decile = case_when(
      between(round(sum_rain_season, 1), round(GS_decile_table[1,2],1) ,round(GS_decile_table[1,3],1) ) ~ "decile_1",
      between(round(sum_rain_season, 1), round(GS_decile_table[2,2],1) ,round(GS_decile_table[2,3],1) ) ~ "decile_2",
      between(round(sum_rain_season, 1), round(GS_decile_table[3,2],1) ,round(GS_decile_table[3,3],1) ) ~ "decile_3",
      between(round(sum_rain_season, 1), round(GS_decile_table[4,2],1) ,round(GS_decile_table[4,3],1) ) ~ "decile_4",
      between(round(sum_rain_season, 1), round(GS_decile_table[5,2],1) ,round(GS_decile_table[5,3],1) ) ~ "decile_5",
      between(round(sum_rain_season, 1), round(GS_decile_table[6,2],1) ,round(GS_decile_table[6,3],1) ) ~ "decile_6",
      between(round(sum_rain_season, 1), round(GS_decile_table[7,2],1) ,round(GS_decile_table[7,3],1) ) ~ "decile_7",
      between(round(sum_rain_season, 1), round(GS_decile_table[8,2],1) ,round(GS_decile_table[8,3],1) ) ~ "decile_8",
      between(round(sum_rain_season, 1), round(GS_decile_table[9,2],1) ,round(GS_decile_table[9,3],1) ) ~ "decile_9",
      between(round(sum_rain_season, 1), round(GS_decile_table[10,2],1) ,round(GS_decile_table[10,3],1) ) ~ "decile_10",
      
      TRUE                      ~ "other"
    )
  )


gs_rain <- gs_rain %>% 
  mutate(Station_name = paste0(distinct(df, Name)),
         Station_number = paste0(distinct(df, Number)) ,
         Years_included = paste0(year(start) ," to ",  year(end)),
         GS_period =   GS_defined_as,
         summer_period =   summer_defined_as 
  )

gs_rain <- gs_rain %>%  rename(gs_sum_rain = sum_rain_season,
                               gs_decile = decile) %>%   select(- season)
summer_rain <- summer_rain %>%
  mutate(
    decile = case_when(
      between(round(sum_rain_season, 1), round(summer_decile_table[1,2],1) ,round(summer_decile_table[1,3],1) ) ~ "decile_1",
      between(round(sum_rain_season, 1), round(summer_decile_table[2,2],1) ,round(summer_decile_table[2,3],1) ) ~ "decile_2",
      between(round(sum_rain_season, 1), round(summer_decile_table[3,2],1) ,round(summer_decile_table[3,3],1) ) ~ "decile_3",
      between(round(sum_rain_season, 1), round(summer_decile_table[4,2],1) ,round(summer_decile_table[4,3],1) ) ~ "decile_4",
      between(round(sum_rain_season, 1), round(summer_decile_table[5,2],1) ,round(summer_decile_table[5,3],1) ) ~ "decile_5",
      between(round(sum_rain_season, 1), round(summer_decile_table[6,2],1) ,round(summer_decile_table[6,3],1) ) ~ "decile_6",
      between(round(sum_rain_season, 1), round(summer_decile_table[7,2],1) ,round(summer_decile_table[7,3],1) ) ~ "decile_7",
      between(round(sum_rain_season, 1), round(summer_decile_table[8,2],1) ,round(summer_decile_table[8,3],1) ) ~ "decile_8",
      between(round(sum_rain_season, 1), round(summer_decile_table[9,2],1) ,round(summer_decile_table[9,3],1) ) ~ "decile_9",
      between(round(sum_rain_season, 1), round(summer_decile_table[10,2],1) ,round(summer_decile_table[10,3],1) ) ~ "decile_10",
      
      TRUE                      ~ "other"
    )
  )


summer_rain <- summer_rain %>%  rename(summer_sum_rain = sum_rain_season,
                                       summer_decile = decile) %>% 
  select(- season)

Year_with_Deciles <- left_join(gs_rain, summer_rain)
Year_with_Deciles <- Year_with_Deciles %>% select(
  year,
  Station_name ,
  Station_number,
  gs_decile,
  gs_sum_rain,
  summer_decile ,
  summer_sum_rain,
  GS_period ,
 summer_period,
  Years_included )


rm(gs_rain, summer_rain)

DT::datatable(Year_with_Deciles)
```

## Save outputs 
* Summer decile table
* GS decile table
* Year with decile info and rainfall info per year and season type (GS, summer)
* Plot of rainfall sum per year and season type
* Daily met file with site info and season type but all date (df_all_date)

"\\fs1-cbr.nexus.csiro.au\{af-sandysoils-ii}\work\Output-1\0.Site-info\Climate\Decile_tables"




```{r save files, echo=FALSE, message=FALSE, warning=FALSE}

Files_saved <- "H:/Output-1/0.Site-info/Climate/Decile_tables/"
               

#decile tables
# GS_decile_table_df
# summer_decile_table_df

write_csv(GS_decile_table_df, 
          file =paste0(Files_saved, "GS_decile_table_df_", site_selected, ".csv"))
write_csv(summer_decile_table_df, 
          file =paste0(Files_saved, "summer_decile_table_df_", site_selected, ".csv"))

#Year_with_Deciles
write_csv(Year_with_Deciles, 
          file =paste0(Files_saved, "Year_with_Deciles_", site_selected, ".csv"))

#plot
 ggsave(plot = plot1,
       filename = paste0(Files_saved,"plot1_sum_rain_per_yr_summer_GS", site_selected, ".png"),
       width = 20, height = 12, units = "cm")

#df (Daily met file with site info and season type)
write_csv(df_all_date, 
          file =paste0(Files_saved, "Daily_met_file_season types_", site_selected, ".csv"))

```

```{r files in directory, echo=FALSE, message=FALSE, warning=FALSE}
Files_saved <- "H:/Output-1/0.Site-info/Climate/Decile_tables/"
files <- list.files(path = Files_saved,full.names = TRUE, pattern = ".csv")

file_info <- file.info(files)
file_info$file_name <- rownames(file_info)
file_info <- file_info %>%
  dplyr::select(atime)
#names(file_info)
DT::datatable(file_info)
```